\section{Introduction}
\label{sec:introduction}

Proportional-Integral-Derivative (PID) controllers remain the dominant
feedback control structure in industrial process control, accounting for more
than 90\% of deployed control loops~\cite{Aastrom2006}.
Despite their conceptual simplicity, choosing appropriate PID parameters for
complex or poorly characterised plants continues to pose a practical challenge
to control engineers.

Classical tuning methods --- such as Ziegler--Nichols (ZN)~\cite{ZieglerNichols1942},
Cohen--Coon~\cite{CohenCoon1953}, and Internal Model Control
(IMC)~\cite{Rivera1986} --- require either an explicit step response
experiment, knowledge of the ultimate gain, or a pre-identified process model.
In industrial settings, performing open-loop step tests is often undesirable
due to production constraints, safety considerations, or the time required to
reach steady state.

Two-point (Bang-Bang) controllers, on the other hand, are already widely
deployed as simple thermostats and safety limiters in cooling, heating, and
refrigeration systems.
These controllers inherently drive the plant into a limit-cycle oscillation
that encodes rich information about the plant dynamics.

\subsection{Motivation}
\label{sec:motivation}

This paper proposes the \emph{Kiwitsche 2-Punkt-Verfahren} (KZV), a procedure
that exploits the Bang-Bang limit cycle to identify the plant and synthesise
a PID controller --- all without requiring an additional open-loop experiment.
The key insight is that the limit cycle produced by a Bang-Bang controller
with known switching levels is equivalent, in the describing-function sense,
to the critical oscillation used in the Ziegler--Nichols closed-loop method.
By additionally recovering the dead time from the phase of the oscillation,
a complete FOPDT model can be estimated, enabling more robust IMC-style
tuning.

\subsection{Industrial Relevance: Cooling Systems}
\label{sec:cooling}

Industrial cooling systems (chillers, refrigeration circuits, climate control)
are a prime application.
Their thermal dynamics are well approximated by second-order-plus-dead-time
(SOPDT) or FOPDT models, and they are routinely operated with Bang-Bang
control during commissioning before a more sophisticated controller is
installed.
The KZV allows this commissioning phase itself to be used for parameter
identification, shortening start-up time.

\subsection{Contributions}
\label{sec:contributions}

The main contributions of this work are:
\begin{enumerate}
    \item A closed-form derivation of FOPDT parameters from Bang-Bang limit
          cycle measurements (\cref{sec:theory}).
    \item Modified IMC-based PID tuning formulas adapted to the KZV
          identification (\cref{sec:methodology}).
    \item Simulation validation on a second-order thermal plant with dead time,
          representative of industrial cooling applications (\cref{sec:results}).
    \item A discussion of robustness to measurement noise and model mismatch
          (\cref{sec:discussion}).
\end{enumerate}

\subsection{Paper Organisation}

\Cref{sec:theory} reviews the theoretical foundations: describing function
analysis of Bang-Bang control and FOPDT modelling.
\Cref{sec:methodology} details the step-by-step KZV procedure.
\Cref{sec:results} presents simulation results.
\Cref{sec:discussion} discusses limitations and future work.
\Cref{sec:conclusion} concludes the paper.
